---
title: "Exponential Smoothing"
author: "Ahmad Al-Kashef"
date: "January 19, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, warning = FALSE, message = FALSE, results = "hold"}
rm(list = ls()) # clear all environment variables
suppressPackageStartupMessages(library(plotly))
```

## Introduction

This is an R Markdown presentation that features a plot created with Plotly.

The subject of the presentation is exponential smoothing, applied to a habit scoring scenario.

You can learn more about it at:

- [Exponential Smoothing Explained](http://www.inventoryops.com/articles/exponential_smoothing.htm])
- [Exponential Smoothing - Wikipedia](https://en.m.wikipedia.org/wiki/Exponential_smoothing])

## Habit Forming Data

The following chunck of R code creates a list of random values of 0 and 1. 

Assuming this is how one person grew a daily habit. The 0's signify the days he missed, and the 1's signify the days he did not.

```{r, warning = FALSE, message = FALSE, results = "hold", echo = TRUE}
set.seed(1)
habit <- c(rbinom(30, 1, 0.05), # 30 values, prob. of 1 = 0.05
           rbinom(60, 1, 0.1), # prob. of 0 = 1 - 0.1
           rbinom(30, 1, 0.5), 
           rbinom(60, 1, 0.7), 
           rbinom(60, 1, 0.8), 
           rbinom(60, 1, 0.9), 
           rbinom(65, 1, 0.25),
           rbinom(30, 1, 0.05),
           rbinom(60, 1, 0.8))
```

## Moving Average

The simplist way to calculate scores of how well (frequent) the habit is kept is by averaging over a weekly window.  

```{r, warning = FALSE, message = FALSE, results = "hold", echo = TRUE}
movingAverage <- rep(0, length(habit))
windowSize <- 7
for (i in 1:(length(habit)-windowSize))
{
  if (i == 1)
  {
    movingAverage[1] <- 0
  } else if ((i %% 7) != 0) {
    movingAverage[i] <- movingAverage[i-1]
  } else {
    movingAverage[i] <- sum(habit[i:(i+windowSize)])/windowSize
  }
}
```

## Exponential Smoothing

The disadvantage of a moving average is that the score is only limited to the window. It doesn't matter how well the habit was maintained before the window. Moving averages also suffer from latency; the size of the window.

Exponential smoothing is aleviates these disadvantages. All the history matters and there is a latency of 2 day only. 

```{r, warning = FALSE, message = FALSE, results = "hold", echo = TRUE}
score1 <- rep(0, length(habit)) # all zeros score list
alpha1 = 0.2
for (i in 3:length(score1)) 
{
 score1[i] <- alpha1 * habit[i-1] + (1 - alpha1) * score1[i-2]
 score1[i] <- round(score1[i], 2)
}
```

## Alpha

*Alpha* is the smoothing factor. It balances the weights between the current state `(habit[i-1])` and the history `score1[i-2]`; how much history you want to take into consideration.

Let's decrease the smoothing factor:

```{r, warning = FALSE, message = FALSE, results = "hold", echo = TRUE}
score2 <- rep(0, length(habit))
alpha2 = 0.02
for (i in 3:length(score2))
{
  score2[i] <- alpha2 * habit[i-1] + (1 - alpha2) * score2[i-2] 
  score2[i] <- round(score2[i], 2)
}
```

## Plotly

The `plotly` library is used to build an interactive plot of the moving average and exponential smoothing scores.

```{r, warning = FALSE, message = FALSE, results = "hold", echo = TRUE}
movingAverageFrame <- data.frame(
  value = movingAverage, 
  index = rep("Moving Average", length(movingAverage)), 
  id = 1:length(movingAverage))
score1Frame <- data.frame(
  value = score1, 
  index = rep("Exp' Smoothing, alpha=0.2", length(score1)), 
  id = 1:length(score1))
score2Frame <- data.frame(
  value = score2, 
  index = rep("Exp' Smoothing, alpha=0.02", length(score2)), 
  id = 1:length(score2))
plotFrame <- rbind(movingAverageFrame, score1Frame, score2Frame)
```

##
```{r, warning = FALSE, message = FALSE, results = "hold", echo = TRUE}
plot_ly() %>% 
  add_lines(x = plotFrame$id, 
            y = plotFrame$value, 
            color = plotFrame$index) %>%
  layout(legend = list(x = 0, y = 1))
```
